---
title: 'STEM Data Analysis: Data Cleaning & Pooling Ordinal and Factor Analysis'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Cleaning, Pooling Ordinal Questions
```{r}
stem_org = read.csv("stem.csv")

# subset out original dataset
stem = stem_org[1:866,]
```

```{r}
# ordinal pooling, questions Q3.1 to Q3.24
barplot(table(stem$Q3.1))
table(stem$Q3.2)

# change to numeric 1-5
reordered = as.data.frame(sapply(stem, function(x) as.integer(factor(as.character(x), levels=c("Strongly Disagree","Disagree","Not Sure", "Agree", "Strongly Agree")))))

stem[24:47] = reordered[24:47]
```

```{r}
# make a copy incase we need original rows for later analysis
stem_ordinal = stem[24:47]
stem_ordinal = na.omit(stem_ordinal)
```

# Pooling Ordinal Questions 3.1 - 3.24
## POOLING RULES: 
- With 5 levels, the smallest quantity must be at least 10% of the data, otherwise combine @ tail.
- after combining 2 columns, if they do not take up at least 15% of the data, then consider combining the 3 closest columns
- do not combine 3 columns if they take up over 50% of entire data.
- after combining 3 columns, if not at least 20% of data, combine again.
- minimum of 2 levels, after pooling, maximum of 5 (no change).

### 3.1 - 3.6
```{r}
#### Q3.1 ####

# checking
barplot(table(stem_ordinal$Q3.1))
cat("Total of all levels:", sum(table(stem_ordinal$Q3.1)), "\n")
cat("Total of levels:", sum(table(stem_ordinal$Q3.1)[1:3]), "\n")

# changing levels
stem_ordinal$Q3.1[stem_ordinal$Q3.1 == 1 | stem_ordinal$Q3.1 == 2 | stem_ordinal$Q3.1 == 3 ] <- 1
stem_ordinal$Q3.1[stem_ordinal$Q3.1 == 4] <- 2
stem_ordinal$Q3.1[stem_ordinal$Q3.1 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.1)) # good

#### Q3.2 ####

# checking
barplot(table(stem_ordinal$Q3.2))
total=sum(table(stem_ordinal$Q3.2))
levels=sum(table(stem_ordinal$Q3.2)[1:2])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")


# changing levels
stem_ordinal$Q3.2[stem_ordinal$Q3.2 == 1 | stem_ordinal$Q3.2 == 2] <- 1
stem_ordinal$Q3.2[stem_ordinal$Q3.2 == 3] <- 2
stem_ordinal$Q3.2[stem_ordinal$Q3.2 == 4] <- 3
stem_ordinal$Q3.2[stem_ordinal$Q3.2 == 5] <- 4


# checking after 
barplot(table(stem_ordinal$Q3.2)) # good

#### Q3.3 ####

# checking
barplot(table(stem_ordinal$Q3.3))
total=sum(table(stem_ordinal$Q3.3))
levels=sum(table(stem_ordinal$Q3.3)[4:5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# changing levels
stem_ordinal$Q3.3[stem_ordinal$Q3.3 == 4 | stem_ordinal$Q3.3 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.3)) # good


#### Q3.4 ####

# checking
barplot(table(stem_ordinal$Q3.4))
total=sum(table(stem_ordinal$Q3.4))
levels=sum(table(stem_ordinal$Q3.4)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # over 10% for 5 levels so clear

# checking after 
barplot(table(stem_ordinal$Q3.4)) # good


#### Q3.5 ####

# checking
barplot(table(stem_ordinal$Q3.5))
total=sum(table(stem_ordinal$Q3.5))
levels=sum(table(stem_ordinal$Q3.5)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# checking after 
barplot(table(stem_ordinal$Q3.5)) # good


#### Q3.6 ####

# checking
barplot(table(stem_ordinal$Q3.6))
total=sum(table(stem_ordinal$Q3.6))
levels=sum(table(stem_ordinal$Q3.6)[1])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # proportion is not at least 10%, combine 1 & 2

# changing levels
stem_ordinal$Q3.6[stem_ordinal$Q3.6 == 1 | stem_ordinal$Q3.3 == 2] <- 1
stem_ordinal$Q3.6[stem_ordinal$Q3.6 == 3] <- 2
stem_ordinal$Q3.6[stem_ordinal$Q3.6 == 4] <- 3
stem_ordinal$Q3.6[stem_ordinal$Q3.6 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.6)) # good
```


### 3.7 - 3.12
```{r}
#### Q3.7 ####

# checking
barplot(table(stem_ordinal$Q3.7))
total=sum(table(stem_ordinal$Q3.7))
levels=sum(table(stem_ordinal$Q3.7)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # proportion is not at least 10%, combine 4 & 5

# changing levels
stem_ordinal$Q3.7[stem_ordinal$Q3.7 == 4 | stem_ordinal$Q3.7 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.7)) # good


#### Q3.8 ####
# checking
barplot(table(stem_ordinal$Q3.8))
total=sum(table(stem_ordinal$Q3.8))
levels=sum(table(stem_ordinal$Q3.8)[1])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # not 10%, combine 1 & 2

# changing levels
stem_ordinal$Q3.8[stem_ordinal$Q3.8 == 1 | stem_ordinal$Q3.8 == 2] <- 1
stem_ordinal$Q3.8[stem_ordinal$Q3.8 == 3] <- 2
stem_ordinal$Q3.8[stem_ordinal$Q3.8 == 4] <- 3
stem_ordinal$Q3.8[stem_ordinal$Q3.8 == 5] <- 4
# checking after 
barplot(table(stem_ordinal$Q3.8)) # good


#### Q3.9 ####
# checking
barplot(table(stem_ordinal$Q3.9))
total=sum(table(stem_ordinal$Q3.9))
levels=sum(table(stem_ordinal$Q3.9)[1])
levels2=sum(table(stem_ordinal$Q3.9)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # factor 1 not 10%, combine 1 & 2. 
cat("Proportion of levels to total", levels2/total, "\n") # Factor 5 not 10%, combine 4 & 5.

# changing levels
stem_ordinal$Q3.9[stem_ordinal$Q3.9 == 1 | stem_ordinal$Q3.9 == 2] <- 1
stem_ordinal$Q3.9[stem_ordinal$Q3.9 == 3] <- 2
stem_ordinal$Q3.9[stem_ordinal$Q3.9 == 4 | stem_ordinal$Q3.9 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.9)) # good


#### Q3.10 ####
# checking
barplot(table(stem_ordinal$Q3.10))
total=sum(table(stem_ordinal$Q3.10))
levels=sum(table(stem_ordinal$Q3.10)[1:3])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# changing levels
stem_ordinal$Q3.10[stem_ordinal$Q3.10 == 1 | stem_ordinal$Q3.10 == 2 |stem_ordinal$Q3.10 == 3] <- 1
stem_ordinal$Q3.10[stem_ordinal$Q3.10 == 4] <- 2
stem_ordinal$Q3.10[stem_ordinal$Q3.10 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.10)) # good


#### Q3.11 ####
# checking
barplot(table(stem_ordinal$Q3.11))
total=sum(table(stem_ordinal$Q3.11))
levels=sum(table(stem_ordinal$Q3.11)[4:5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # factors 4 & 5 are 20%, combine.

# changing levels
stem_ordinal$Q3.11[stem_ordinal$Q3.11 == 4 | stem_ordinal$Q3.11 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.11)) # good


#### Q3.12 ####
# checking
barplot(table(stem_ordinal$Q3.12))
total=sum(table(stem_ordinal$Q3.12))
levels=sum(table(stem_ordinal$Q3.12)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # smallest factor takes up over 10%, no changes.

# changing levels
# no changing levels

# checking after
barplot(table(stem_ordinal$Q3.12)) # good
```
### 3.13 - 3.18
```{r}
#### Q3.13 ####
# checking
barplot(table(stem_ordinal$Q3.13))
total=sum(table(stem_ordinal$Q3.13))
levels=sum(table(stem_ordinal$Q3.13)[1])
levels2=sum(table(stem_ordinal$Q3.13)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 
cat("Proportion of levels to total", levels2/total, "\n") 
# level5 by itself not over 10%, combine with 4. level 1 by itself not ove 10%, combine with 2.

# changing levels
stem_ordinal$Q3.13[stem_ordinal$Q3.13 == 1 | stem_ordinal$Q3.13 == 2] <- 1
stem_ordinal$Q3.13[stem_ordinal$Q3.13 == 3] <- 2
stem_ordinal$Q3.13[stem_ordinal$Q3.13 == 4 | stem_ordinal$Q3.13 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.13))


#### Q3.14 ####
# checking
barplot(table(stem_ordinal$Q3.14))
total=sum(table(stem_ordinal$Q3.14))
levels=sum(table(stem_ordinal$Q3.14)[1])
levels2=sum(table(stem_ordinal$Q3.14)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 
cat("Proportion of levels to total", levels2/total, "\n") 
# combine level 1 & 2, since less than 10%

# changing levels
stem_ordinal$Q3.14[stem_ordinal$Q3.14 == 1 | stem_ordinal$Q3.14 == 2] <- 1
stem_ordinal$Q3.14[stem_ordinal$Q3.14 == 3] <- 2
stem_ordinal$Q3.14[stem_ordinal$Q3.14 == 4] <- 3
stem_ordinal$Q3.14[stem_ordinal$Q3.14 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.14))


#### Q3.15 ####
barplot(table(stem_ordinal$Q3.15))

total=sum(table(stem_ordinal$Q3.15))
levels=sum(table(stem_ordinal$Q3.15)[1:3])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# changing levels
stem_ordinal$Q3.15[stem_ordinal$Q3.15 == 1 | stem_ordinal$Q3.15 == 2 | stem_ordinal$Q3.15 == 3] <- 1
stem_ordinal$Q3.15[stem_ordinal$Q3.15 == 4] <- 2
stem_ordinal$Q3.15[stem_ordinal$Q3.15 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.15))


#### Q3.16 ####
# checking
barplot(table(stem_ordinal$Q3.16))
total=sum(table(stem_ordinal$Q3.16))
levels=sum(table(stem_ordinal$Q3.16)[4:5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# changing levels
stem_ordinal$Q3.16[stem_ordinal$Q3.16 == 1] <- 1
stem_ordinal$Q3.16[stem_ordinal$Q3.16 == 2] <- 2
stem_ordinal$Q3.16[stem_ordinal$Q3.16 == 3] <- 3
stem_ordinal$Q3.16[stem_ordinal$Q3.16 == 4| stem_ordinal$Q3.16 == 5] <- 4
# checking after 
barplot(table(stem_ordinal$Q3.16))


#### Q3.17 ####
# checking
barplot(table(stem_ordinal$Q3.17))
total=sum(table(stem_ordinal$Q3.17))
levels=sum(table(stem_ordinal$Q3.17)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # smallest factor, 5, not over 10% on its own so combine with 4.


# changing levels
stem_ordinal$Q3.17[stem_ordinal$Q3.17 == 1] <- 1
stem_ordinal$Q3.17[stem_ordinal$Q3.17 == 2] <- 2
stem_ordinal$Q3.17[stem_ordinal$Q3.17 == 3] <- 3
stem_ordinal$Q3.17[stem_ordinal$Q3.17 == 4| stem_ordinal$Q3.17 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.17))


#### Q3.18 ####
# checking
barplot(table(stem_ordinal$Q3.18))
total=sum(table(stem_ordinal$Q3.18))
levels=sum(table(stem_ordinal$Q3.18)[1:2])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.18[stem_ordinal$Q3.18 == 1| stem_ordinal$Q3.18 == 2] <- 1
stem_ordinal$Q3.18[stem_ordinal$Q3.18 == 3] <- 2
stem_ordinal$Q3.18[stem_ordinal$Q3.18 == 4] <- 3
stem_ordinal$Q3.18[stem_ordinal$Q3.18 == 5] <- 4
# checking after 
barplot(table(stem_ordinal$Q3.18))
```

### 3.19 - 3.24

```{r}
#### Q3.19 ####
# checking
barplot(table(stem_ordinal$Q3.19))
total=sum(table(stem_ordinal$Q3.19))
levels=sum(table(stem_ordinal$Q3.19)[1:3])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.19[stem_ordinal$Q3.19 == 1 | stem_ordinal$Q3.19 == 2 | stem_ordinal$Q3.19 == 3 ] <- 1
stem_ordinal$Q3.19[stem_ordinal$Q3.19 == 4] <- 2
stem_ordinal$Q3.19[stem_ordinal$Q3.19 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.19))


#### Q3.20 ####
# checking
barplot(table(stem_ordinal$Q3.20))
total=sum(table(stem_ordinal$Q3.20))
levels=sum(table(stem_ordinal$Q3.20)[1:3])
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.20[stem_ordinal$Q3.20 == 1 | stem_ordinal$Q3.20 == 2 | stem_ordinal$Q3.20 == 3] <- 1
stem_ordinal$Q3.20[stem_ordinal$Q3.20 == 4] <- 2
stem_ordinal$Q3.20[stem_ordinal$Q3.20 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.20))



#### Q3.21 ####
# checking
barplot(table(stem_ordinal$Q3.21))
total=sum(table(stem_ordinal$Q3.21))
levels=sum(table(stem_ordinal$Q3.21)[1:3]) 
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 
# after combining 3 levels, still under 10% of total, so we combine 4.

# changing levels
stem_ordinal$Q3.21[stem_ordinal$Q3.21 == 1 | stem_ordinal$Q3.21 == 2 | stem_ordinal$Q3.21 == 3 |  stem_ordinal$Q3.21 == 4] <- 1
stem_ordinal$Q3.21[stem_ordinal$Q3.21 == 5] <- 2

# checking after 
barplot(table(stem_ordinal$Q3.21))


#### Q3.22 ####
# checking
barplot(table(stem_ordinal$Q3.22))
total=sum(table(stem_ordinal$Q3.22))
levels=sum(table(stem_ordinal$Q3.22)[1:3]) 
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.22[stem_ordinal$Q3.22 == 1 | stem_ordinal$Q3.22 == 2 | stem_ordinal$Q3.22 == 3] <- 1
stem_ordinal$Q3.22[stem_ordinal$Q3.22 == 4] <- 2
stem_ordinal$Q3.22[stem_ordinal$Q3.22 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.22))


#### Q3.23 ####
# checking
barplot(table(stem_ordinal$Q3.23))
total=sum(table(stem_ordinal$Q3.23))
levels=sum(table(stem_ordinal$Q3.23)[1:3]) 
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.23[stem_ordinal$Q3.23 == 1 | stem_ordinal$Q3.23 == 2 | stem_ordinal$Q3.23 == 3] <- 1
stem_ordinal$Q3.23[stem_ordinal$Q3.23 == 4] <- 2
stem_ordinal$Q3.23[stem_ordinal$Q3.23 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.23))


#### Q3.24 ####
# checking
barplot(table(stem_ordinal$Q3.24))
total=sum(table(stem_ordinal$Q3.24))
levels=sum(table(stem_ordinal$Q3.24)[1:3]) 
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.24[stem_ordinal$Q3.24 == 1 | stem_ordinal$Q3.24 == 2 | stem_ordinal$Q3.24 == 3] <- 1
stem_ordinal$Q3.24[stem_ordinal$Q3.24 == 4] <- 2
stem_ordinal$Q3.24[stem_ordinal$Q3.24 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.24))
```

```{r}
# note: the new file has 821 rows instead of the original 866, because we removed NA's for pooling.
stem2 = stem
stem2_filtered = stem2[rownames(stem2) %in% rownames(stem_ordinal),]
stem2_filtered[24:47] = stem_ordinal

pooled_stem = stem2_filtered
```

# Factor Analysis
```{r}
library(psych)
library(GPArotation)
library(psychTools)
stem = pooled_stem
```

```{r}
# EDA, Correlations
stem_ordinal = stem[24:47]
lowerCor(stem_ordinal)
corPlot(stem_ordinal)

# pick number of factors 
# The blue line shows eigenvalues of actual data and the two red lines (placed on top of each other) show simulated and resampled data. Here we look at the large drops in the actual data and spot the point where it levels off to the right. Also we locate the point of inflection – the point where the gap between simulated data and actual data tends to be minimum.
parallel = fa.parallel(stem_ordinal, fa='fa', fm='ml')
parallel$fa.values

# eigenvalues over 1 = 2 factors
# eigenvalues over 0.7 = 3 factors
```
```{r}
# only consider the loadings more than 0.3 and not loading on more than one factor (called double loading). Note that negative values are acceptable here. 

factor_analysis <- function(n, rotate_method) {
  factor <- fa(stem_ordinal,nfactors = n,rotate = rotate_method, fm="ml")
  cat(n, "Factors")
  print(factor$loadings,cutoff = 0.3)
  
  return(factor)
}  


# factors
twof = factor_analysis(2, "promax")
threef = factor_analysis(3, "promax")

anova(twof, threef) # significantly different
```
```{r}
threef
# checking adequacy
# rmsr = 0.05
# rmsea = 0.073 adequate
# tli = 0.746 not good
# cfi = 1 - ((threef$STATISTIC - threef$dof)/( threef$null.chisq - threef$null.dof)) = 0.8 not good

# reliability
alpha(stem_ordinal[,c(15,16,19,20,21,22,23,24)], check.keys=TRUE) # good, over 0.7 & 0.8
alpha(stem_ordinal[,c(8,9,10,13,18)],check.keys=TRUE) # 0.69 alpha
alpha(stem_ordinal[,c(4,5,6)]) # 0.68 alpha
```


```{r}
### making new variables 
# SenseBelonging = 3.15, 3.16 (inverse this), 3.19 - 3.24
# Stress = 3.8, 3.9, 3.10, 3.13 (inverse this),  3.18
# SES = 3.4, 3.5, 3.6

rescale <- function(x_i,n){
    n-x_i
}

# check rescaling
table(stem$Q3.16)
table(sapply(stem$Q3.16,rescale,5))

table(stem$Q3.13)
table(sapply(stem$Q3.13,rescale,4))

# making variables
stem$Belonging_Factor = stem$Q3.15 + sapply(stem$Q3.16,rescale,5) + stem$Q3.19 + stem$Q3.20 + stem$Q3.21 + stem$Q3.22 + stem$Q3.23 + stem$Q3.24

stem$Stress_Factor = stem$Q3.8  + stem$Q3.9 + stem$Q3.10 + sapply(stem$Q3.13,rescale,4) + stem$Q3.18

stem$SES_Factor = stem$Q3.4 + stem$Q3.5 + stem$Q3.6

# save as csv for further analysis & answer research questions
write.csv(stem, "stem_newfactors.csv", row.names=FALSE)
```

