---
title: "STEM Data Analysis: Data Cleaning & Pooling Ordinal Data"
output: html_notebook
---

```{r}
stem_org = read.csv("stem.csv")

# subset out original dataset
stem = stem_org[1:866,]
```

```{r}
# ordinal pooling, questions Q3.1 to Q3.24
barplot(table(stem$Q3.1))
table(stem$Q3.2)

# change to numeric 1-5
reordered = as.data.frame(sapply(stem, function(x) as.integer(factor(as.character(x), levels=c("Strongly Disagree","Disagree","Not Sure", "Agree", "Strongly Agree")))))

stem[24:47] = reordered[24:47]
```

```{r}
# make a copy incase we need original rows for later analysis
stem_ordinal = stem[24:47]
stem_ordinal = na.omit(stem_ordinal)
```

# Pooling Ordinal Questions 3.1 - 3.24
## POOLING RULES: 
- With 5 levels, the smallest quantity must be at least 10% of the data, otherwise combine @ tail.
- after combining 2 columns, if they do not take up at least 15% of the data, then consider combining the 3 closest columns
- do not combine 3 columns if they take up over 50% of entire data.
- after combining 3 columns, if not at least 20% of data, combine again.
- minimum of 2 levels, after pooling, maximum of 5 (no change).

### 3.1 - 3.6
```{r}
#### Q3.1 ####

# checking
barplot(table(stem_ordinal$Q3.1))
cat("Total of all levels:", sum(table(stem_ordinal$Q3.1)), "\n")
cat("Total of levels:", sum(table(stem_ordinal$Q3.1)[1:3]), "\n")

# changing levels
stem_ordinal$Q3.1[stem_ordinal$Q3.1 == 1 | stem_ordinal$Q3.1 == 2 | stem_ordinal$Q3.1 == 3 ] <- 1
stem_ordinal$Q3.1[stem_ordinal$Q3.1 == 4] <- 2
stem_ordinal$Q3.1[stem_ordinal$Q3.1 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.1)) # good

#### Q3.2 ####

# checking
barplot(table(stem_ordinal$Q3.2))
total=sum(table(stem_ordinal$Q3.2))
levels=sum(table(stem_ordinal$Q3.2)[1:2])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")


# changing levels
stem_ordinal$Q3.2[stem_ordinal$Q3.2 == 1 | stem_ordinal$Q3.2 == 2] <- 1
stem_ordinal$Q3.2[stem_ordinal$Q3.2 == 3] <- 2
stem_ordinal$Q3.2[stem_ordinal$Q3.2 == 4] <- 3
stem_ordinal$Q3.2[stem_ordinal$Q3.2 == 5] <- 4


# checking after 
barplot(table(stem_ordinal$Q3.2)) # good

#### Q3.3 ####

# checking
barplot(table(stem_ordinal$Q3.3))
total=sum(table(stem_ordinal$Q3.3))
levels=sum(table(stem_ordinal$Q3.3)[4:5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# changing levels
stem_ordinal$Q3.3[stem_ordinal$Q3.3 == 4 | stem_ordinal$Q3.3 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.3)) # good


#### Q3.4 ####

# checking
barplot(table(stem_ordinal$Q3.4))
total=sum(table(stem_ordinal$Q3.4))
levels=sum(table(stem_ordinal$Q3.4)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # over 10% for 5 levels so clear

# checking after 
barplot(table(stem_ordinal$Q3.4)) # good


#### Q3.5 ####

# checking
barplot(table(stem_ordinal$Q3.5))
total=sum(table(stem_ordinal$Q3.5))
levels=sum(table(stem_ordinal$Q3.5)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# checking after 
barplot(table(stem_ordinal$Q3.5)) # good


#### Q3.6 ####

# checking
barplot(table(stem_ordinal$Q3.6))
total=sum(table(stem_ordinal$Q3.6))
levels=sum(table(stem_ordinal$Q3.6)[1])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # proportion is not at least 10%, combine 1 & 2

# changing levels
stem_ordinal$Q3.6[stem_ordinal$Q3.6 == 1 | stem_ordinal$Q3.3 == 2] <- 1
stem_ordinal$Q3.6[stem_ordinal$Q3.6 == 3] <- 2
stem_ordinal$Q3.6[stem_ordinal$Q3.6 == 4] <- 3
stem_ordinal$Q3.6[stem_ordinal$Q3.6 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.6)) # good
```


### 3.7 - 3.12
```{r}
#### Q3.7 ####

# checking
barplot(table(stem_ordinal$Q3.7))
total=sum(table(stem_ordinal$Q3.7))
levels=sum(table(stem_ordinal$Q3.7)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # proportion is not at least 10%, combine 4 & 5

# changing levels
stem_ordinal$Q3.7[stem_ordinal$Q3.7 == 4 | stem_ordinal$Q3.7 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.7)) # good


#### Q3.8 ####
# checking
barplot(table(stem_ordinal$Q3.8))
total=sum(table(stem_ordinal$Q3.8))
levels=sum(table(stem_ordinal$Q3.8)[1])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # not 10%, combine 1 & 2

# changing levels
stem_ordinal$Q3.8[stem_ordinal$Q3.8 == 1 | stem_ordinal$Q3.8 == 2] <- 1
stem_ordinal$Q3.8[stem_ordinal$Q3.8 == 3] <- 2
stem_ordinal$Q3.8[stem_ordinal$Q3.8 == 4] <- 3
stem_ordinal$Q3.8[stem_ordinal$Q3.8 == 5] <- 4
# checking after 
barplot(table(stem_ordinal$Q3.8)) # good


#### Q3.9 ####
# checking
barplot(table(stem_ordinal$Q3.9))
total=sum(table(stem_ordinal$Q3.9))
levels=sum(table(stem_ordinal$Q3.9)[1])
levels2=sum(table(stem_ordinal$Q3.9)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # factor 1 not 10%, combine 1 & 2. 
cat("Proportion of levels to total", levels2/total, "\n") # Factor 5 not 10%, combine 4 & 5.

# changing levels
stem_ordinal$Q3.9[stem_ordinal$Q3.9 == 1 | stem_ordinal$Q3.9 == 2] <- 1
stem_ordinal$Q3.9[stem_ordinal$Q3.9 == 3] <- 2
stem_ordinal$Q3.9[stem_ordinal$Q3.9 == 4 | stem_ordinal$Q3.9 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.9)) # good


#### Q3.10 ####
# checking
barplot(table(stem_ordinal$Q3.10))
total=sum(table(stem_ordinal$Q3.10))
levels=sum(table(stem_ordinal$Q3.10)[1:3])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# changing levels
stem_ordinal$Q3.10[stem_ordinal$Q3.10 == 1 | stem_ordinal$Q3.10 == 2 |stem_ordinal$Q3.10 == 3] <- 1
stem_ordinal$Q3.10[stem_ordinal$Q3.10 == 4] <- 2
stem_ordinal$Q3.10[stem_ordinal$Q3.10 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.10)) # good


#### Q3.11 ####
# checking
barplot(table(stem_ordinal$Q3.11))
total=sum(table(stem_ordinal$Q3.11))
levels=sum(table(stem_ordinal$Q3.11)[4:5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # factors 4 & 5 are 20%, combine.

# changing levels
stem_ordinal$Q3.11[stem_ordinal$Q3.11 == 4 | stem_ordinal$Q3.11 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.11)) # good


#### Q3.12 ####
# checking
barplot(table(stem_ordinal$Q3.12))
total=sum(table(stem_ordinal$Q3.12))
levels=sum(table(stem_ordinal$Q3.12)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # smallest factor takes up over 10%, no changes.

# changing levels
# no changing levels

# checking after
barplot(table(stem_ordinal$Q3.12)) # good
```
### 3.13 - 3.18
```{r}
#### Q3.13 ####
# checking
barplot(table(stem_ordinal$Q3.13))
total=sum(table(stem_ordinal$Q3.13))
levels=sum(table(stem_ordinal$Q3.13)[1])
levels2=sum(table(stem_ordinal$Q3.13)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 
cat("Proportion of levels to total", levels2/total, "\n") 
# level5 by itself not over 10%, combine with 4. level 1 by itself not ove 10%, combine with 2.

# changing levels
stem_ordinal$Q3.13[stem_ordinal$Q3.13 == 1 | stem_ordinal$Q3.13 == 2] <- 1
stem_ordinal$Q3.13[stem_ordinal$Q3.13 == 3] <- 2
stem_ordinal$Q3.13[stem_ordinal$Q3.13 == 4 | stem_ordinal$Q3.13 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.13))


#### Q3.14 ####
# checking
barplot(table(stem_ordinal$Q3.14))
total=sum(table(stem_ordinal$Q3.14))
levels=sum(table(stem_ordinal$Q3.14)[1])
levels2=sum(table(stem_ordinal$Q3.14)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 
cat("Proportion of levels to total", levels2/total, "\n") 
# combine level 1 & 2, since less than 10%

# changing levels
stem_ordinal$Q3.14[stem_ordinal$Q3.14 == 1 | stem_ordinal$Q3.14 == 2] <- 1
stem_ordinal$Q3.14[stem_ordinal$Q3.14 == 3] <- 2
stem_ordinal$Q3.14[stem_ordinal$Q3.14 == 4] <- 3
stem_ordinal$Q3.14[stem_ordinal$Q3.14 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.14))


#### Q3.15 ####
barplot(table(stem_ordinal$Q3.15))

total=sum(table(stem_ordinal$Q3.15))
levels=sum(table(stem_ordinal$Q3.15)[1:3])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# changing levels
stem_ordinal$Q3.15[stem_ordinal$Q3.15 == 1 | stem_ordinal$Q3.15 == 2 | stem_ordinal$Q3.15 == 3] <- 1
stem_ordinal$Q3.15[stem_ordinal$Q3.15 == 4] <- 2
stem_ordinal$Q3.15[stem_ordinal$Q3.15 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.15))


#### Q3.16 ####
# checking
barplot(table(stem_ordinal$Q3.16))
total=sum(table(stem_ordinal$Q3.16))
levels=sum(table(stem_ordinal$Q3.16)[4:5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n")

# changing levels
stem_ordinal$Q3.16[stem_ordinal$Q3.16 == 1] <- 1
stem_ordinal$Q3.16[stem_ordinal$Q3.16 == 2] <- 2
stem_ordinal$Q3.16[stem_ordinal$Q3.16 == 3] <- 3
stem_ordinal$Q3.16[stem_ordinal$Q3.16 == 4| stem_ordinal$Q3.16 == 5] <- 4
# checking after 
barplot(table(stem_ordinal$Q3.16))


#### Q3.17 ####
# checking
barplot(table(stem_ordinal$Q3.17))
total=sum(table(stem_ordinal$Q3.17))
levels=sum(table(stem_ordinal$Q3.17)[5])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") # smallest factor, 5, not over 10% on its own so combine with 4.


# changing levels
stem_ordinal$Q3.17[stem_ordinal$Q3.17 == 1] <- 1
stem_ordinal$Q3.17[stem_ordinal$Q3.17 == 2] <- 2
stem_ordinal$Q3.17[stem_ordinal$Q3.17 == 3] <- 3
stem_ordinal$Q3.17[stem_ordinal$Q3.17 == 4| stem_ordinal$Q3.17 == 5] <- 4

# checking after 
barplot(table(stem_ordinal$Q3.17))


#### Q3.18 ####
# checking
barplot(table(stem_ordinal$Q3.18))
total=sum(table(stem_ordinal$Q3.18))
levels=sum(table(stem_ordinal$Q3.18)[1:2])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.18[stem_ordinal$Q3.18 == 1| stem_ordinal$Q3.18 == 2] <- 1
stem_ordinal$Q3.18[stem_ordinal$Q3.18 == 3] <- 2
stem_ordinal$Q3.18[stem_ordinal$Q3.18 == 4] <- 3
stem_ordinal$Q3.18[stem_ordinal$Q3.18 == 5] <- 4
# checking after 
barplot(table(stem_ordinal$Q3.18))
```

### 3.19 - 3.24

```{r}
#### Q3.19 ####
# checking
barplot(table(stem_ordinal$Q3.19))
total=sum(table(stem_ordinal$Q3.19))
levels=sum(table(stem_ordinal$Q3.19)[1:3])

cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.19[stem_ordinal$Q3.19 == 1 | stem_ordinal$Q3.19 == 2 | stem_ordinal$Q3.19 == 3 ] <- 1
stem_ordinal$Q3.19[stem_ordinal$Q3.19 == 4] <- 2
stem_ordinal$Q3.19[stem_ordinal$Q3.19 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.19))


#### Q3.20 ####
# checking
barplot(table(stem_ordinal$Q3.20))
total=sum(table(stem_ordinal$Q3.20))
levels=sum(table(stem_ordinal$Q3.20)[1:3])
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.20[stem_ordinal$Q3.20 == 1 | stem_ordinal$Q3.20 == 2 | stem_ordinal$Q3.20 == 3] <- 1
stem_ordinal$Q3.20[stem_ordinal$Q3.20 == 4] <- 2
stem_ordinal$Q3.20[stem_ordinal$Q3.20 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.20))



#### Q3.21 ####
# checking
barplot(table(stem_ordinal$Q3.21))
total=sum(table(stem_ordinal$Q3.21))
levels=sum(table(stem_ordinal$Q3.21)[1:3]) 
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 
# after combining 3 levels, still under 10% of total, so we combine 4.

# changing levels
stem_ordinal$Q3.21[stem_ordinal$Q3.21 == 1 | stem_ordinal$Q3.21 == 2 | stem_ordinal$Q3.21 == 3 |  stem_ordinal$Q3.21 == 4] <- 1
stem_ordinal$Q3.21[stem_ordinal$Q3.21 == 5] <- 2

# checking after 
barplot(table(stem_ordinal$Q3.21))


#### Q3.22 ####
# checking
barplot(table(stem_ordinal$Q3.22))
total=sum(table(stem_ordinal$Q3.22))
levels=sum(table(stem_ordinal$Q3.22)[1:3]) 
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.22[stem_ordinal$Q3.22 == 1 | stem_ordinal$Q3.22 == 2 | stem_ordinal$Q3.22 == 3] <- 1
stem_ordinal$Q3.22[stem_ordinal$Q3.22 == 4] <- 2
stem_ordinal$Q3.22[stem_ordinal$Q3.22 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.22))


#### Q3.23 ####
# checking
barplot(table(stem_ordinal$Q3.23))
total=sum(table(stem_ordinal$Q3.23))
levels=sum(table(stem_ordinal$Q3.23)[1:3]) 
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.23[stem_ordinal$Q3.23 == 1 | stem_ordinal$Q3.23 == 2 | stem_ordinal$Q3.23 == 3] <- 1
stem_ordinal$Q3.23[stem_ordinal$Q3.23 == 4] <- 2
stem_ordinal$Q3.23[stem_ordinal$Q3.23 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.23))


#### Q3.24 ####
# checking
barplot(table(stem_ordinal$Q3.24))
total=sum(table(stem_ordinal$Q3.24))
levels=sum(table(stem_ordinal$Q3.24)[1:3]) 
cat("Total of levels:", levels, "\n")
cat("Proportion of levels to total", levels/total, "\n") 

# changing levels
stem_ordinal$Q3.24[stem_ordinal$Q3.24 == 1 | stem_ordinal$Q3.24 == 2 | stem_ordinal$Q3.24 == 3] <- 1
stem_ordinal$Q3.24[stem_ordinal$Q3.24 == 4] <- 2
stem_ordinal$Q3.24[stem_ordinal$Q3.24 == 5] <- 3

# checking after 
barplot(table(stem_ordinal$Q3.24))
```

```{r}
# note: the new file has 821 rows instead of the original 866, because we removed NA's for pooling.
stem2 = stem
stem2_filtered = stem2[rownames(stem2) %in% rownames(stem_ordinal),]
stem2_filtered[24:47] = stem_ordinal

# save csv, use for factor analysis + other analysis
write.csv(stem2_filtered, "pooled_stem.csv")
```









