---
title: "Stress Analysis"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

```{r}
stem = read.csv("stem_newfactors.csv", na.strings=c("",NA))
head(stem)
```

### Exploratory Analysis
```{r}
library(gdata)

# clean data
stem$Gender[stem$Gender == 'Genderfluid' |  stem$Gender == 'Transgender female'] <- NA
stem = stem[!is.na(stem$Gender),]
stem$Orientation[stem$Orientation == 'straight'] <- 'Heterosexual'
stem = drop.levels(stem)


# gender
barplot(tapply(stem$Stress_Factor, stem$Gender, mean), main="Gender vs. Average Stress Level")

# orientation
barplot(tapply(stem$Stress_Factor, stem$Orientation, mean), main="Sexual Orientation vs. Average Stress Level")

# year
# refactor order
stem$Year <- factor(stem$Year, levels = c("Freshman", "Sophomore", "Junior", "Senior", "Fifth Year", "Graduate Student"))
barplot(tapply(stem$Stress_Factor, stem$Year, mean), main="Year vs. Average Stress Level")


# international
barplot(tapply(stem$Stress_Factor, stem$International, mean,na.rm=TRUE), main="International vs. Average Stress Level")

# transfer
barplot(tapply(stem$Stress_Factor, stem$Transfer, mean,na.rm=TRUE), main="Transfer vs. Average Stress Level")

# efl
barplot(tapply(stem$Stress_Factor, stem$EFL, mean), main="EFL vs. Average Stress Level")

# first gen
barplot(tapply(stem$Stress_Factor, stem$FirstGen, mean,na.rm=TRUE), main="First Gen vs. Average Stress Level")

# gpa
stem$GPA[stem$GPA == 0] <- NA
#range(stem$GPA,na.rm=TRUE)

stem$GPA[stem$GPA < 3.0] = "< 3.0"
stem$GPA[stem$GPA >= 3.0 & stem$GPA <= 3.5] = "3.0 <= x <= 3.5"
stem$GPA[stem$GPA > 3.5] = ">3.5"

stem$GPA <- factor(stem$GPA, levels = c("< 3.0", "3.0 <= x <= 3.5", ">3.5"))

barplot(tapply(stem$Stress_Factor, stem$GPA, mean,na.rm=TRUE), main="GPA vs. Average Stress Level")
```

```{r}
# q3.2, .5, .15, year, gender
# how correlated are these variables with the stress_factor variable

stem$GPA[stem$GPA == 0] = mean(stem$GPA) 
target = stem[,'Stress_Factor']

q2 = stem[,'Q3.2']
q4 = stem[,'Q3.4']
q5 = stem[,'Q3.5']
q6 = stem[,'Q3.6']
q15 = stem[,'Q3.15']
q14 = stem[,'Q3.14']
ses = stem[,'SES_Factor']


q2a<- aov(Stress_Factor~as.factor(q2), data = stem)
summary(q2a) #significant
barplot(tapply(target, q2, mean), main = 'Q3.2 versus Stress Level', ylab = 'Stress Level', xlab = 'Level of preparedness from hs for STEM') 

q4a<- aov(Stress_Factor~as.factor(q4), data = stem)
summary(q4a) # significant
barplot(tapply(target, q4, mean), main = 'Q3.4 versus Stress Level', ylab = 'Stress Level', xlab = 'Family struggled financially growing up') 

q5a<- aov(Stress_Factor~as.factor(q5), data = stem)
summary(q5a) # significant
barplot(tapply(target, q5, mean), main = 'Q3.5 versus Stress Level', ylab = 'Stress Level', xlab = 'Level of work needed to pay for expenses') 

q6a<- aov(Stress_Factor~as.factor(q6), data = stem)
summary(q6a) # significant
barplot(tapply(target, q6, mean), main = 'Q3.6 versus Stress Level', ylab = 'Stress Level', xlab = 'Level graduating and needing  to find a job immediately') 

q14a<- aov(Stress_Factor~as.factor(q14), data = stem)
summary(q14a) # significant
barplot(tapply(target, q14, mean), main = 'Q3.14 versus Stress Level', ylab = 'Stress Level', xlab = 'To what extent do you seek help when you have an academic problem')

# correlation for numeric vars
cor(ses, target)
res <- cor.test(ses, target, 
                    method = "pearson")
res$p.value
barplot(tapply(target, ses, mean), main = 'SES versus Stress Level', ylab = 'Stress Level', xlab = 'SES', ylim = c(8,12))

# q3.2, 3.4, 3.5, 3.6, 3.14, SES are all significant
```

```{r}
# exploratory analysis: belonging questions (taken from factor analysis) 3.15, 3.16, 3.19 - 3.24 checking significance with ANOVA
q15 = stem[,'Q3.15']
q16 = stem[,'Q3.16']
q19 = stem[,'Q3.19']
q20 = stem[,'Q3.20']
q21 = stem[,'Q3.21']
q22 = stem[,'Q3.22']
q23 = stem[,'Q3.23']
q24 = stem[,'Q3.24']

q15a<- aov(Stress_Factor~as.factor(q15), data = stem)
summary(q15a) # not significant
barplot(tapply(target, q15, mean), main = 'Q3.15 versus Stress Level', ylab = 'Stress Level', xlab = 'To what extent do you know what you want to do after graduation') 


q16a<- aov(Stress_Factor~as.factor(q16), data = stem)
summary(q16a) # significant
barplot(tapply(target, q16, mean), main = 'Q3.16 versus Stress Level', ylab = 'Stress Level', xlab = 'Q3.16')

q19a<- aov(Stress_Factor~as.factor(q19), data = stem)
summary(q19a) # significant
barplot(tapply(target, q19, mean), main = 'Q3.19 versus Stress Level', ylab = 'Stress Level', xlab = 'Q3.19')

q20a<- aov(Stress_Factor~as.factor(q20), data = stem)
summary(q20a) # significant
barplot(tapply(target, q20, mean), main = 'Q3.20 versus Stress Level', ylab = 'Stress Level', xlab = 'Q3.20')

q21a<- aov(Stress_Factor~as.factor(q21), data = stem)
summary(q21a) # significant
barplot(tapply(target, q21, mean), main = 'Q3.21 versus Stress Level', ylab = 'Stress Level', xlab = 'Q3.21')

q22a<- aov(Stress_Factor~as.factor(q22), data = stem)
summary(q22a) # significant
barplot(tapply(target, q22, mean), main = 'Q3.22 versus Stress Level', ylab = 'Stress Level', xlab = 'Q3.22')

q23a<- aov(Stress_Factor~as.factor(q23), data = stem)
summary(q23a) # significant
barplot(tapply(target, q23, mean), main = 'Q3.23 versus Stress Level', ylab = 'Stress Level', xlab = 'Q3.23')

q24a<- aov(Stress_Factor~as.factor(q24), data = stem)
summary(q24a) # significant
barplot(tapply(target, q24, mean), main = 'Q3.24 versus Stress Level', ylab = 'Stress Level', xlab = 'Q3.24')

# Q3.16, 3.19, 3.20 - 3.24 all significant
```


```{r}
# ANOVA
gender <- aov(Stress_Factor~Gender, data = stem)
summary(gender) # significantly different

orientation <- aov(Stress_Factor~Orientation, data = stem)
summary(orientation) # not significantly different

year <- aov(Stress_Factor~Year, data = stem)
summary(year) # not significantly different

international <- aov(Stress_Factor~International, data = stem)
summary(international) # significantly different

transfer <- aov(Stress_Factor~Transfer, data = stem)
summary(transfer) # not significantly different

efl <- aov(Stress_Factor~EFL, data = stem)
summary(efl) # not significantly different

firstgen <- aov(Stress_Factor~FirstGen, data = stem)
summary(firstgen) # significantly different

# simple linear reg
gpa_lm <- lm(Stress_Factor ~ GPA, data = stem)
summary(gpa_lm) # gpa is significant 

# we will use gender, firstgen, and gpa in our linear reg, Q3.2, Q3.5, Q3.14
# q3.2, 3.4, 3.5, 3.6, 3.14, SES are all significant
# Q3.16, 3.19, 3.20 - 3.24 all significant
```

## Modeling
We have broken up our final variables into 3 separate subcategories, or blocks. We should check each block individually and see which is most important.

Block1: SES - Q3.2, Q3.4, Q3.5, Q3.6

Block2: Demographics - Gender, FirstGen

Block3: Academic Awareness/Direction: Q3.14, GPA

Block4: Belonging Q3.16, 3.19, 3.20-3.24
```{r}
# block1
ses = lm(Stress_Factor ~ as.factor(Q3.2) + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6), data=stem)
summary(ses)
# adj R2 0.1255

# block 2 
dem = lm(Stress_Factor ~ Gender + FirstGen, data=stem)
summary(dem)
# adj R2 0.06056

# block 3 
aca = lm(Stress_Factor ~ as.factor(Q3.14) + GPA, data=stem)
summary(aca)
# adj R2 0.06923

# block 4 
bel = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23)+ as.factor(Q3.24), data=stem)
summary(bel)
# adj R2 0.2324. use this as base 

# interaction 
interaction = lm(Stress_Factor ~ as.factor(Q3.2) * GPA, data=stem)
summary(interaction)
# adj R2 0.1056 use this as base 


# block4 + block1
m1 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.2) + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6), data=stem)
summary(m1)
# adj R2 0.2789, highest of 2 blocks

# block4 + block2
m2 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + Gender + FirstGen, data=stem)
summary(m2)
# adj R2 0.2574


# block4 + block3
m3 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + GPA, data=stem)
summary(m3)
# adj R2 0.266

# block4 + block3 + block1
m4 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + GPA + as.factor(Q3.2) + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6), data=stem)
summary(m4)
# adj R2 0.2984 highest

# block4 + block3 + block2
m5 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + GPA + Gender + FirstGen, data=stem)
summary(m5)
# adj R2 0.2898

# block4 + block3 + block2 + block1
m6 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + GPA + Gender + FirstGen + as.factor(Q3.2) + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6), data=stem)
summary(m6)
# adj R2 0.3163

# block4 + block3 + block2 + block1 + interaction
m7 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + Gender + FirstGen + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6) + as.factor(Q3.2) * GPA, data=stem)
summary(m7)
# adj R2 0.326
```
Block4: Belonging had the most influence on increasing R2 for Stress.  

## Plots
```{r}
stem$Q3.2 = as.factor(stem$Q3.2)

interaction.plot(stem$GPA, stem$Q3.2, stem$Stress_Factor,xlab="GPA", ylab="Stress_Factor", trace.label="HS Prep STEM")
```

For students who strongly felt their HS prepped them well for STEM (level=4), their stress levels decreased drastically as GPA increases.
Overall students who strongly felt their HS did NOT prep for STEM (level=1) had the highest overall stress, across all GPAs, and their stress levels only slightly decrease with higher GPA. 
For students who felt their HS slightly prepared them for STEM (level=2), had high stress at low GPAs and dropped drastically as GPA increased. 
For students who felt their HS prepared them for STEM (level=3) had the overall lowest stress throughout all GPAs, and stress levels remained the same no matter what GPA the students had.






