---
title: "Stress Analysis"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

```{r}
stem = read.csv("stem_newfactors.csv", na.strings=c("",NA))
head(stem)
```

### Exploratory Analysis
```{r}
library(gdata)

# clean data
stem$Gender[stem$Gender == 'Genderfluid' |  stem$Gender == 'Transgender female'] <- NA
stem = stem[!is.na(stem$Gender),]
stem$Orientation[stem$Orientation == 'straight'] <- 'Heterosexual'
stem = drop.levels(stem)

# make a function for exploratory plotting
plotting <- function(xvar_name, xvar, title, xlab, ylab) {
  df <- data.frame(xvar_name = levels(factor(xvar)), Stress_Factor = tapply(stem$Stress_Factor, xvar, mean))
  ggplot(na.omit(df)) +
  aes(x=xvar_name, y=Stress_Factor) + 
  labs(x = xlab,
       y = ylab,
       title = title,
       fill = xlab) +
  geom_bar(stat="identity", aes(fill = xvar_name)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
}  

# Gender
# Levels: Female, Male
plotting(Gender, stem$Gender, "Gender vs. Average Stress Level", "Gender", "Stress_Factor")


# Sexual Orientation
# Levels: Bisexual, Gay/ Homosexual, Heterosexual, Pansexual, Queer, I prefer not to say
plotting(Orientation, stem$Orientation, "Sexual Orientation vs. Average Stress Level", "Sexual Orientation", "Stress_Factor")

# Year in College
# Levels: Freshman, Sophomore, Junior, Senior, Fifth Year, Graduate Student
df <- data.frame(Year= levels(factor(stem$Year)), Stress_Factor = tapply(stem$Stress_Factor, stem$Year, mean))
df$Year <- factor(df$Year, levels = c("Freshman", "Sophomore", "Junior", "Senior", "Fifth Year", "Graduate Student")) # refactoring order to reflect standard higher education progression in America

ggplot(na.omit(df)) +
  aes(x=Year, y=Stress_Factor, color=Year) + 
  labs(x = "Year in College",
       y = "Stress_Factor",
       title = "Year in College vs. Average Stress Level") +
  geom_bar(stat="identity", aes(fill = Year)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# International 
# Levels: Yes, No (Question: Are you an international student?)
plotting(International, stem$International, "International vs. Average Stress Level", "International Student?", "Stress_Factor")


# Transfer
# Levels: Yes, No (Question: Did you transfer from a community college?)
plotting(Transfer, stem$Transfer, "Transfer vs. Average Stress Level", "Transfer Student?", "Stress_Factor")


# EFL (English as a Foreign Language)
# Levels: Yes, No (Question: Is English your first language?)
plotting(EFL, stem$EFL, "EFL vs. Average Stress Level", "EFL?", "Stress_Factor")

df <- data.frame(EFL= levels(factor(stem$EFL)), Stress_Factor = tapply(stem$Stress_Factor, stem$EFL, mean))

# First Generation
# Levels: Yes, No (Question: Are you the first generation in your family to attend college?)
plotting(FirstGen, stem$FirstGen, "First Gen vs. Average Stress Level", "First Generation?", "Stress_Factor")



# GPA
stem$GPA[stem$GPA == 0] <- NA
#range(stem$GPA,na.rm=TRUE)

stem$GPA[stem$GPA < 3.0] = "< 3.0"
stem$GPA[stem$GPA >= 3.0 & stem$GPA <= 3.5] = "3.0 <= x <= 3.5"
stem$GPA[stem$GPA > 3.5] = ">3.5"

stem$GPA <- factor(stem$GPA, levels = c("< 3.0", "3.0 <= x <= 3.5", ">3.5"))

df <- data.frame(GPA= levels(factor(stem$GPA)), Stress_Factor = tapply(stem$Stress_Factor, stem$GPA, mean))
df$GPA <- factor(df$GPA, levels = c("< 3.0", "3.0 <= x <= 3.5", ">3.5"))

ggplot(na.omit(df)) +
  aes(x=GPA, y=Stress_Factor, color=GPA) + 
  labs(x = "GPA",
       y = "Stress_Factor",
       title = "GPA vs. Average Stress Level") +
  geom_bar(stat="identity", aes(fill = GPA)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# exploring SES questions: q3.2, 4, 5, 6, 14, 15, ses_factor

stem$GPA[stem$GPA == 0] = mean(stem$GPA) 
target = stem[,'Stress_Factor']

q2 = stem[,'Q3.2']
q4 = stem[,'Q3.4']
q5 = stem[,'Q3.5']
q6 = stem[,'Q3.6']
q15 = stem[,'Q3.15']
q14 = stem[,'Q3.14']
ses = stem[,'SES_Factor']


q2a<- aov(Stress_Factor~as.factor(q2), data = stem)
summary(q2a) #significant
plotting(Q3.2, stem$Q3.2,"Q3.2 versus Avg. Stress Level","Level of preparedness from hs for STEM", "Stress Level")


q4a<- aov(Stress_Factor~as.factor(q4), data = stem)
summary(q4a) # significant
plotting(Q3.4, stem$Q3.4,"Q3.4 versus Avg. Stress Level","Family struggled financially growing up", "Stress Level")


q5a<- aov(Stress_Factor~as.factor(q5), data = stem)
summary(q5a) # significant
plotting(Q3.5, stem$Q3.5,"Q3.5 versus Avg. Stress Level","Level of work needed to pay for expenses", "Stress Level")


q6a<- aov(Stress_Factor~as.factor(q6), data = stem)
summary(q6a) # significant
plotting(Q3.6, stem$Q3.6,"Q3.6 versus Avg. Stress Level","Level graduating and needing  to find a job immediately", "Stress Level")


q14a<- aov(Stress_Factor~as.factor(q14), data = stem)
summary(q14a) # significant
plotting(Q3.14, stem$Q3.14,"Q3.14 versus Avg. Stress Level","Seek help with academic problem", "Stress Level")

# correlation for numeric vars
cor(ses, target)
res <- cor.test(ses, target, 
                    method = "pearson")
res$p.value

# q3.2, 3.4, 3.5, 3.6, 3.14, SES are all significant
```

```{r}
# exploratory analysis: belonging questions (taken from factor analysis) 3.15, 3.16, 3.19 - 3.24 checking significance with ANOVA
q15 = stem[,'Q3.15']
q16 = stem[,'Q3.16']
q19 = stem[,'Q3.19']
q20 = stem[,'Q3.20']
q21 = stem[,'Q3.21']
q22 = stem[,'Q3.22']
q23 = stem[,'Q3.23']
q24 = stem[,'Q3.24']

q15a<- aov(Stress_Factor~as.factor(q15), data = stem)
summary(q15a) # not significant
plotting(Q3.15, stem$Q3.15,"Q3.15 versus Avg. Stress Level","Know what to do after grad", "Stress Level")


q16a<- aov(Stress_Factor~as.factor(q16), data = stem)
summary(q16a) # significant
plotting(Q3.16, stem$Q3.16,"Q3.16 versus Avg. Stress Level","Feel excluded at UCLA", "Stress Level")


q19a<- aov(Stress_Factor~as.factor(q19), data = stem)
summary(q19a) # significant
plotting(Q3.19, stem$Q3.19,"Q3.19 versus Avg. Stress Level","Feel belonging at UCLA", "Stress Level")


q20a<- aov(Stress_Factor~as.factor(q20), data = stem)
summary(q20a) # significant
plotting(Q3.20, stem$Q3.20,"Q3.20 versus Avg. Stress Level","Parents understand UCLA standards", "Stress Level")


q21a<- aov(Stress_Factor~as.factor(q21), data = stem)
summary(q21a) # significant
plotting(Q3.21, stem$Q3.21,"Q3.21 versus Avg. Stress Level","Parents value college education", "Stress Level")


q22a<- aov(Stress_Factor~as.factor(q22), data = stem)
summary(q22a) # significant
plotting(Q3.22, stem$Q3.22,"Q3.22 versus Avg. Stress Level","Happy at UCLA", "Stress Level")


q23a<- aov(Stress_Factor~as.factor(q23), data = stem)
summary(q23a) # significant
plotting(Q3.23, stem$Q3.23,"Q3.23 versus Avg. Stress Level","Confidence in ability to succeed at UCLA", "Stress Level")


q24a<- aov(Stress_Factor~as.factor(q24), data = stem)
summary(q24a) # significant
plotting(Q3.24, stem$Q3.24,"Q3.24 versus Avg. Stress Level","Diverse group interaction at UCLA", "Stress Level")

# Q3.16, 3.19, 3.20 - 3.24 all significant
```


```{r}
# ANOVA
gender <- aov(Stress_Factor~Gender, data = stem)
summary(gender) # significantly different

orientation <- aov(Stress_Factor~Orientation, data = stem)
summary(orientation) # not significantly different

year <- aov(Stress_Factor~Year, data = stem)
summary(year) # not significantly different

international <- aov(Stress_Factor~International, data = stem)
summary(international) # significantly different

transfer <- aov(Stress_Factor~Transfer, data = stem)
summary(transfer) # not significantly different

efl <- aov(Stress_Factor~EFL, data = stem)
summary(efl) # not significantly different

firstgen <- aov(Stress_Factor~FirstGen, data = stem)
summary(firstgen) # significantly different

# simple linear reg
gpa_lm <- lm(Stress_Factor ~ GPA, data = stem)
summary(gpa_lm) # gpa is significant 

# we will use gender, firstgen, and gpa in our linear reg, Q3.2, Q3.5, Q3.14
# q3.2, 3.4, 3.5, 3.6, 3.14, SES are all significant
# Q3.16, 3.19, 3.20 - 3.24 all significant
```


```{r}
## Indicators
# Analyzing relationship between SES which is an indicator for socioeconomic struggle with Stress_Factor. It represents the sum of responses to three questions, Q3.4 - My family struggled financially when I was growing up, Q3.5 - I need to work to pay for my expenses and Q3.6 - I need to graduate and get a job as soon as possible; with each question being measured as a continuous variable from 0 to 1 (0 =  Low level of struggle and 1 = High level of struggle). We use significance level = 0.01 for the best results.

cor(ses, target)
res <- cor.test(ses, target, 
                    method = "pearson")
res$p.value # p-value is statistically significant

df <- data.frame(SES= levels(factor(ses)), y = tapply(target, ses, mean))
df$SES <- factor(df$SES, levels = c("3","4","5","6","7","8","9","10","11","12","13","14"))

ggplot(df) +
  aes(x=SES, y=y, color=SES) + 
  labs(x = "SES",
       y = "Stress Level",
       title = "SES versus Stress Level") +
  geom_bar(stat="identity", aes(fill = SES)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# P-value is statistically significant, however, we cannot accurately use SES as a predictor in our model due to unequal sample sizes for each level of the SES_Factor indicator.
```


## Modeling
We have broken up our final variables into 3 separate subcategories, or blocks. We should check each block individually and see which is most important.

Block1: SES - Q3.2, Q3.4, Q3.5, Q3.6

Block2: Demographics - Gender, FirstGen

Block3: Academic Awareness/Direction: Q3.14, GPA

Block4: Belonging Q3.16, 3.19, 3.20-3.24
```{r}
# block1
ses = lm(Stress_Factor ~ as.factor(Q3.2) + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6), data=stem)
summary(ses)
# adj R2 0.1255

# block 2 
dem = lm(Stress_Factor ~ Gender + FirstGen, data=stem)
summary(dem)
# adj R2 0.06056

# block 3 
aca = lm(Stress_Factor ~ as.factor(Q3.14) + GPA, data=stem)
summary(aca)
# adj R2 0.06923

# block 4 
bel = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23)+ as.factor(Q3.24), data=stem)
summary(bel)
# adj R2 0.2324. use this as base 

# interaction 
interaction = lm(Stress_Factor ~ as.factor(Q3.2) * GPA, data=stem)
summary(interaction)
# adj R2 0.1056 use this as base 


# block4 + block1
m1 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.2) + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6), data=stem)
summary(m1)
# adj R2 0.2789, highest of 2 blocks

# block4 + block2
m2 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + Gender + FirstGen, data=stem)
summary(m2)
# adj R2 0.2574


# block4 + block3
m3 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + GPA, data=stem)
summary(m3)
# adj R2 0.266

# block4 + block3 + block1
m4 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + GPA + as.factor(Q3.2) + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6), data=stem)
summary(m4)
# adj R2 0.2984 highest

# block4 + block3 + block2
m5 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + GPA + Gender + FirstGen, data=stem)
summary(m5)
# adj R2 0.2898

# block4 + block3 + block2 + block1
m6 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + GPA + Gender + FirstGen + as.factor(Q3.2) + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6), data=stem)
summary(m6)
# adj R2 0.3163

# block4 + block3 + block2 + block1 + interaction
m7 = lm(Stress_Factor ~ as.factor(Q3.16) + as.factor(Q3.19) + as.factor(Q3.20) + as.factor(Q3.21) + as.factor(Q3.22) + as.factor(Q3.23) + as.factor(Q3.24) + as.factor(Q3.14) + Gender + FirstGen + as.factor(Q3.4) + as.factor(Q3.5) + as.factor(Q3.6) + as.factor(Q3.2) * GPA, data=stem)
summary(m7)
# adj R2 0.326
```
Block4: Belonging had the most influence on increasing R2 for Stress.  

```{r}
summary(m7)
```


## Plots
```{r}
GPA = factor(stem$GPA, levels = c("< 3.0", "3.0 <= x <= 3.5", ">3.5")) 
Q3.2 = as.factor(stem$Q3.2)
Stress_Factor = as.numeric(stem$Stress_Factor)

interaction <- glm(Stress_Factor~GPA*Q3.2, family=gaussian)

interaction.plot(GPA, stem$Q3.2, stem$Stress_Factor,xlab="GPA", ylab="Stress_Factor", trace.label="HS Prep STEM")
plot(allEffects(interaction), xlab="Q3.2 (Level of STEM preparation in High School)", ylab="Stress_Factor", main="Interaction of GPA and HS Prep in STEM on Stress_Factor", ask=FALSE)
```
Overall, stress tends to go down with higher GPA, as students with GPAs below 3.0 had higher stress and students with GPA 3.0 or higher tended to have lower stress. Students with GPA under 3.0 had fluctuating stress levels with increasing preparedness for STEM. There was a drastic drop in stress at HS prep = 3, and a prominent increase in stress at HS prep = 4. 

Overall students who strongly felt their HS did NOT prep for STEM (level=1) had the highest overall stress, across all GPAs, and their stress levels only slightly decrease with higher GPA. Students who felt their HS prepared them for STEM (level=3) had the overall lowest stress throughout all GPAs, and stress levels remained the same no matter what GPA the students had.  
For students who felt their HS slightly prepared them for STEM (level=2), had high stress at low GPAs and dropped drastically as GPA increased. Similarly, students who strongly felt their HS prepped them well for STEM (level=4), their stress levels decreased drastically as GPA increases. 







